---
title: "HW1"
output:
  pdf_document: default
  html_document:
    df_print: paged
---
Due at noon, Thursday September 6, 2018.
Information on how to submit the HW (softcopy) will be given on moodle. 

We will use data from the gapminder package, let's load it. 
```{r setup}
library(tidyverse)
#install.packages("gapminder")
library(gapminder)
```

A look at the data:
```{r}
gapminder
```

A simple base R plot of life expectancy against log(GDP):
```{r}
# Base R plot
plot(lifeExp ~ log(gdpPercap), data = gapminder)
```

# HW exercises
Goal: gain experience with ggplot and reading help files/using google searches to get help on R functions

1. make one scatter plot using ggplot with
- Life expectancy on the y-axis, label axis as ``Life expectancy''
- GDP per capita on the x-axis, use a log-transform of the x-axis while still displaying the (unlogged) GDP values 
(update per moodle: fine to use a log10 transform)
- shapes to indicate the continent, add a legend
- color to indicate the year, do NOT add a legend for that


```{r}
ggplot(gapminder,
       aes(x = (gdpPercap), y = lifeExp)) +
  geom_point(aes(shape = continent)) +
  geom_point(aes(shape = continent, color = year), show.legend=F) +
  labs(y = "Life expectancy") +
  scale_x_log10() # here i use base 10 for the log transform 
```
2. add two geom_smooth layers (a) and (b) to the plot you made in (1) where
(a) is fitted to the entire dataset and uses default settings for the smoother used and plotting settings
(b) is fitted to data in Africa only, provides a 99% confidence interval, is plotted in red (both point estimates as well as confidence interval), and is transparent (so if it overlaps with (a), (a) is still visible). 
Hint: to fit to data in Africa only, consider using "data = filter(gapminder, continent == "Africa")"

Use the help function (or the internet) to figure out what argument to add where!

```{r}
ggplot(gapminder,
       aes(x = (gdpPercap), y = lifeExp)) +
  geom_point(aes(shape = continent)) +
  geom_point(aes(shape = continent, color = year), show.legend=F) +
  labs(y = "Life expectancy") +
  scale_x_log10() +
  geom_smooth() +
  geom_smooth(data = filter(gapminder, continent == "Africa"), alpha = 0.2, 
              level = 0.99, col = 2, fill = "red") 
```
