---
title: "HW3: Transforming and tidying prevalence data"
output:
  html_document:
    df_print: paged
  pdf_document: default
---
Due on Friday September 28, 2018, to submit via github using this link:

https://classroom.github.com/a/BZpoQMcd

Note that the last exercise 3 is based on material that we haven't yet finished in class (hence the Fri deadline); we will go through it relatively quickly on Tu.


The data set prevdat (defined in the set-up chunk below) contains information on prevalence of a disease on 10 populations from 1980 until 2017, stored as follows:

* pop = index of the population, 10 unique ones in total. The coding is pop1 through pop10, combined with info on a grouping (A or B)

* year = observation year

* indicator = prevalence or population 

* value = gives pop size when indicator == population, prevalence (in %) otherwise
NOTE: if prevalence is missing for a population-year, this means that it is equal to zero. 

# Exercise:

1. per year, calculate prevalence for the 10 populations combined, taking into account the varying population sizes across populations. Store your output as a tibble called "exercise1" with columns year and tot_prevalence.
Note: you will need to take into account that missing population-prevalence-years have zero prevalence. 
2. Calculate prevalence for groupings A and B and create a line graph that shows the prevalence over time for both groupings.

3. Not based on prevdat:
Using gss_cat, summarize mean tv hours by marital group as in the graph as shown on the next page. 


```{r setup}
library(tidyverse)
# read in the data using this link (you need to be connected to the internet)
prevdat <- read_rds(gzcon(url("https://people.umass.edu/lalkema/prevdat_hw3.rds")))
```



```{r}
#Exercise1
 exercise1 <- prevdat %>%
  spread(indicator,value) %>%
  drop_na(prevalence) %>%
  group_by(year) %>%
  summarise(tot_prevalence=sum(popsize*prevalence)/sum(popsize))
 #group_by(pop,year) %>%
  #summarise(tot=sum(popsize),tot_prevalence=sum(prevalence*popsize)/tot) %>%
  #ungroup() %>%
  #select(year,tot_prevalence)
 


#ggplot(exercise1)+geom_line(aes(year,tot_prevalence))
 
```

```{r}
#Exercise2
 totA <- prevdat %>%
  spread(indicator,value) %>%
  drop_na(prevalence) %>%
  mutate(newline= str_sub(pop,-1)) %>%
spread(newline,pop) %>%
  group_by(year,A) %>%
  drop_na(A) %>%
  summarise(totA_prevalence=sum(popsize*prevalence)/sum(popsize)) %>%
  group_by(year) %>%
  summarise(totA_prevalence = mean(totA_prevalence))
  
  
totB <- prevdat %>%
  spread(indicator,value) %>%
  drop_na(prevalence) %>%
  mutate(newline= str_sub(pop,-1)) %>%
spread(newline,pop) %>%
  group_by(year,B) %>%
  drop_na(B) %>%
  summarise(totB_prevalence=sum(popsize*prevalence)/sum(popsize)) %>%
  group_by(year) %>%
  summarise(totB_prevalence=mean(totB_prevalence))
  
 
  
tot <- totA %>%
left_join(totB,by = "year") 
  ggplot(tot)+geom_line(aes(year,totA_prevalence),color="blue")+geom_line(aes(year,totB_prevalence),color="red")+ylab("GroupA(BLUE)--GroupB(RED)")
```




```{r}
#Exercise3
gss_cat %>%
  drop_na(tvhours) %>%
  mutate(marital = marital %>%
           fct_collapse( Married=c("Married","Widowed","Separated"),Notmarried=c("Divorced","Never married") ) %>%
  fct_lump("No answer")
  ) %>%
  group_by(marital) %>%
  summarise(tvhours=mean(tvhours))  %>%
 ggplot(aes(x =fct_reorder(marital,tvhours,desc=FALSE) , y = tvhours)) + geom_col() +xlab("Marital status")+coord_flip()

```

